---
title: "Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aggregate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The primary function for aggregation is `aggregate_data()`. You can view the full documentation using `?aggregate_data` or `help("aggregate_data")`. A few key things to note:

* Requires input data in the format described in the 'Getting started' article

* If the `vaccination_data` input is set to `NULL`, the function will ignore your `design_selection` and perform historical data aggregation instead.

* The input datasets are saved and removed from the global environment during aggregation, but restored afterwards by default (see `restore_input_data` parameter). They can also be restored manually using the `recombine_input_data()` function. Once aggregation is complete, the saved versions of the input datasets are deleted. If you would like to store the input datasets that were used, please do so manually whilst they are still in your global environment.

* Validation of the input datasets is run automatically before aggregation is started. If any issues are found, aggregation will be aborted and the issues printed in the console. These issues will need to be fixed before aggregation can be completed. Validation can also be run separately - see Advanced usage section at the end of this document.

* There is no output from the function, however it will create a new folder with the name format "RCA\_[STUDY\_CODENAME]\_[SITE_CODE]\_[CYCLE\_START]-[CYCLE\_END]", at the location specified by `working_directory`, and save the aggregated tables within it. In a multi-site study, this folder should be zipped and shared with the coordinating centre. Output files are written in **Parquet format** by default (`output_format = "parquet"`), which is readable in R via `arrow::read_parquet()`, Python, and SQL engines. Use `output_format = "csv"` for csv output.

* Requires an 'options.txt' file to run - in a multi-site study this is provided by the coordinating centre; in a single-site study you create it yourself. It should be saved in an appropriate location (specified by `options_file_location`), and the same file used for the duration of the study (one options file per study module). For a full description of the file format and every configurable field, see the relevant article or run `vignette("options", "rapidcycler")`.

### Set up for examples

Load the package:

```{r setup}
library(rapidcycler)
```

If you would like to use synthetic data to run the examples, you can generate and load it directly:

```{r, eval=FALSE}
synth <- generate_synthetic_data(pop_size = 100000)
person_data     <- synth$person_data
vaccination_data <- synth$vaccination_data
outcome_data    <- synth$outcome_data
```

By default, `generate_synthetic_data()` also saves `person_data.csv`, `vaccination_data.csv`, and `outcome_data.csv` to the current working directory (`save_data = TRUE`). To load previously saved files:

```{r, eval=FALSE}
person_data      <- data.table::fread("person_data.csv")
vaccination_data <- data.table::fread("vaccination_data.csv")
outcome_data     <- data.table::fread("outcome_data.csv")
```

### Example: Default usage

There are 6 required inputs for the function, and several optional inputs that will revert to a default if not specified.

The code below will create a new folder in your current working directory, given by `getwd()`. There must also be an 'options.txt' file saved in this location.

```{r, eval=FALSE}
# Assuming person_data, vaccination_data, and outcome_data are data.table objects already loaded into the global environment
aggregate_data(person_data, vaccination_data, outcome_data,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE")
# design_selection defaults to NULL, which runs all 5 analytical designs (plus descriptives always)
```

### Example: Study designs

The `design_selection` parameter controls which study designs are aggregated (see manuscript for details). By default (`design_selection = NULL`) all five designs are run. You can pass a character vector of design names to run a specific subset, from the following 5 options:

| `"self_post"` | Self-controlled post-vaccination control window |
| `"self_pre"` | Self-controlled pre-vaccination control window |
| `"historical"` | Cohort design with historic comparator |
| `"concurrent_vac"` | Cohort design with concurrent vaccinated comparator group |
| `"concurrent_unvac"` | Cohort design with concurrent unvaccinated comparator group |

Descriptive outcomes and vaccination datasets are **always produced** for non-historical aggregations and are not controlled by `design_selection`. These contain raw counts of outcomes and vaccinations for the cycle.

```{r, eval=FALSE}
# Run only the self-controlled designs
aggregate_data(person_data, vaccination_data, outcome_data,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               design_selection = c("self_pre", "self_post"))
```

### Example: Aggregation of historical data

The aggregation function can also be used to do a one-off aggregation of historical data, necessary for the cohort design with historic comparator. Simply provide historical person_data and outcome_data datasets to the `aggregate_data()` function, with `vaccination_data = NULL`, and it will aggregate them appropriately. When `vaccination_data = NULL`, the `design_selection` is ignored and two output files are produced instead: `data_historic_cases` (outcome counts by period, age group, sex, AESI, and encounter type) and `data_historic_person_days` (person-days by period, age group, and sex).

```{r, eval=FALSE}
# Aggregate historical data
aggregate_data(person_data, vaccination_data = NULL, outcome_data,
               cycle_start_date = as.Date("2017-01-01"),
               cycle_end_date = as.Date("2024-12-31"),
               site_code = "TEST_SITE")
```

### Example: Using different locations

There are two important file locations within the function:

* `options_file_location`: this is where the 'options.txt' file is stored (in local memory)

* `working_directory`: this is where a new folder will be created, within which aggregated tables are saved

Both these inputs will default to the current working directory, given by `getwd()`, if not specified. To use a different location, you can either change your current working directory using `setwd()` (both locations will be the same), or use these inputs to specify different locations (as in the recommended folder structure). 

```{r, eval=FALSE}
# Assuming person_data, vaccination_data, and outcome_data are data.table objects already loaded into the global environment
aggregate_data(person_data, vaccination_data, outcome_data,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               options_file_location = "~/Documents/RCA/STUDY/SUBSTUDY/",
               working_directory = "~/Documents/RCA/STUDY/SUBSTUDY/data")
```

### Example: Memory considerations

#### Insufficient RAM

If you find that you are running out of RAM during execution of the `aggregate_data()` function, or execution is very slow, you can restrict the amount of data that is aggregated at one time using the `split_size` input. The function splits the input data into portions of size `split_size` according to `PID`, stores the split datasets in a subfolder called 'data', aggregates one split at a time, and then recombines and performs the final aggregation once all splits are complete. The default is 500,000, and should be lowered if memory is an issue. If you have large amounts of RAM, increasing the split size may also speed up aggregation. **NOTE: you should still input the whole datasets - the function will handle the splitting itself**.

```{r, eval=FALSE}
# Assuming person_data, vaccination_data, and outcome_data are data.table objects already loaded into the global environment
aggregate_data(person_data, vaccination_data, outcome_data,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               split_size = 100000)
```


#### Input datasets too large

**NOTE: only use this method if your input datasets are so large that it is not possible to load them into R in one go**. If so, you can divide the data into chunks before aggregating. If you do so, you must specify `input_data_in_chunks = TRUE`. The data should be divided into chunks by `PID`, so that all the records associated with a person are included in the same chunk. The temporary results from each chunk will be stored until the last chunk is aggregated (specified using `final_chunk = TRUE`), after which the temporary results will be combined and the final aggregation performed.

In this case, you will need to call the `aggregate_data()` function multiple times. Each time, you should pass only a single chunk of each dataset, therefore, requires loading a new chunk of input data (either from a saved file or directly from the database) before each function call.

The example below assumes that the input datasets are split into 2 chunks, and there exists code to retrieve these chunks from a saved file or directly from the database.

```{r, eval=FALSE}
person_data_chunk1 <- load_person_data_chunk1()
vaccination_data_chunk1 <- load_vaccination_data_chunk1()
outcome_data_chunk1 <- load_outcome_data_chunk1()

aggregate_data(person_data_chunk1, vaccination_data_chunk1, outcome_data_chunk1,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               input_data_in_chunks = TRUE,
               final_chunk = FALSE) # FALSE until final chunk

# Remove data from memory (can also do this automatically by setting 'restore_input_data = FALSE' in function call)
rm("person_data_chunk1", "vaccination_data_chunk1", "outcome_data_chunk1")

person_data_chunk2 <- load_person_data_chunk2()
vaccination_data_chunk2 <- load_vaccination_data_chunk2()
outcome_data_chunk2 <- load_outcome_data_chunk2()

aggregate_data(person_data_chunk2, vaccination_data_chunk2, outcome_data_chunk2,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               input_data_in_chunks = TRUE,
               final_chunk = TRUE) # TRUE to trigger final aggregation

# Remove data from memory (can also do this automatically by setting 'restore_input_data = FALSE' in function call)
rm("person_data_chunk2", "vaccination_data_chunk2", "outcome_data_chunk2")
```

**IMPORTANT - do not combine fully aggregated tables together, as this will not produce correctly aggregated tables. Dealing with the input data in chunks must be performed in the way described above**

### Example: Stopping and restarting aggregation

It is possible to stop the aggregation and restart at a later time. You can stop aggregation by manually interrupting execution, or by using the `stop_after_n_splits` input. Either way, aggregation can be resumed from where it was stopped. Note that, if you stop aggregation by manually interrupting execution, the input datasets will not be restored to the global environment afterwards. However, they are saved in the 'data' subfolder (in split format) ready for resumption, and can be restored to the global environment using the `recombine_input_data()` function.

The example below will stop aggregation after 2 splits of the input datasets have been aggregated (note, this will aggregate those splits across all time periods and study designs).

```{r, eval=FALSE}
# Assuming person_data, vaccination_data, and outcome_data are data.table objects already loaded into the global environment
aggregate_data(person_data, vaccination_data, outcome_data,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               stop_after_n_splits = 2)
```

The simplest way to resume aggregation is by using the `resume_cycle_aggregation()` function:

```{r, eval=FALSE}
resume_cycle_aggregation(folder_name = "RCA_COVID_TEST_SITE_20250101-20250131", 
                         working_directory = "~/Documents/RCA/STUDY/SUBSTUDY/data")

# NOTE: also optional arguments stop_after_n_splits, input_data_in_chunks, and final_chunk, that behave in the same way as previously described
```


### Output files

When aggregation completes, the results folder contains one file per output dataset plus a `notes.txt` metadata file.

**Non-historical aggregation** (vaccination data provided):

| File | Produced when | Contents |
|:---|:---|:---|
| `data_descriptive_outcomes.parquet` | Always | Outcome counts by period, age group, sex, and encounter type |
| `data_descriptive_vaccinations.parquet` | Always | Vaccination counts by period, age group, sex, vaccine subtype, type, and dose |
| `data_self_post.parquet` | `"self_post"` in `design_selection` | Risk / control counts and window ratios using post-vaccination control window |
| `data_self_pre.parquet` | `"self_pre"` in `design_selection` | Risk / control counts and window ratios using pre-vaccination control window |
| `data_exposed_cases.parquet` | `"historical"` in `design_selection` | Outcome counts in the post-vaccination risk window for vaccinated persons |
| `data_exposed_person_days.parquet` | `"historical"` in `design_selection` | Person-days in the post-vaccination risk window for vaccinated persons |
| `data_concurrent_vac.parquet` | `"concurrent_vac"` in `design_selection` | Risk / control counts and group ratios using vaccinated comparator group |
| `data_concurrent_unvac.parquet` | `"concurrent_unvac"` in `design_selection` | Risk / control counts and group ratios using unvaccinated comparator group |
| `notes.txt` | Always | Metadata: cycle dates, site code, study designs run, options settings, and descriptive statistics |

**Historical aggregation** (`vaccination_data = NULL`):

| File | Contents |
|:---|:---|
| `data_historic_cases.parquet` | Outcome counts by period, age group, sex, AESI, and encounter type |
| `data_historic_person_days.parquet` | Person-days by period, age group, and sex |
| `notes.txt` | Metadata: cycle dates, site code, options settings, and descriptive statistics |

For non-historical aggregation, files excluded via `design_selection` will not be present in the folder.

If `output_format = "csv"` is used, files will have `.csv` extensions instead.

Output files can be read back into R using:

```{r, eval=FALSE}
library(arrow)
results <- arrow::read_parquet("RCA_STUDY_TEST_SITE_20250101-20250131/data_self_post.parquet")
```

### Example: Suppression of small counts

To protect against re-identification, counts below a minimum threshold can be suppressed before the results folder is shared. Use the `suppression_limit` parameter to set this threshold. Any count strictly less than `suppression_limit` (but greater than zero) will be replaced with `-1` in the output.

Note that suppression will significantly hinder the ability to use the aggregated data for sequential analysis, so should be avoided or minimised as much as possible.

Suppression applies to the `COUNT`, `CASE_COUNT`, `CONTROL_COUNT`, and `PERSON_DAYS` columns across all output files where present. Zero counts are never suppressed. Only primary suppression is performed - the ability to reconstruct a low count based on its surrounding cells is not considered.

```{r, eval=FALSE}
aggregate_data(person_data, vaccination_data, outcome_data,
               cycle_start_date = as.Date("2025-01-01"),
               cycle_end_date = as.Date("2025-01-31"),
               site_code = "TEST_SITE",
               suppression_limit = 5) # suppress counts of 1-4
```




