---
title: "The options.txt configuration file"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The options.txt configuration file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `options.txt` file is a plain-text configuration file that controls how `aggregate_data()` runs for your study module. It contains study-level settings that remain fixed for the duration of the study: the analysis period structure, the outcomes of interest, the vaccine codes, and the parameters of each study design.

How you obtain or create your `options.txt` depends on how the package is being used:

- **Multi-site implementation**: the coordinating centre provides the `options.txt` for your study module. You should use the file provided without modification — it encodes the agreed study protocol and ensures all sites use consistent settings. Contact your study coordinator if you have not yet received it.
- **Single-site implementation**: there is no coordinating centre. You are responsible for creating your own `options.txt`. Use the annotated template in this package as a starting point (see below).

This vignette documents the file format and every field, so that you can:

- verify that the file you have received is correct and complete,
- understand how each setting influences the aggregation output, and
- create a valid file from scratch for a single-site study.

A copy of the annotated template can be opened in R at any time:

```{r eval = FALSE}
file.show(system.file("extdata", "options_template.txt", package = "rapidcycler"))
```

---

## File format

`options.txt` is a plain-text file. Each setting occupies one line in `key: value` format:

```
study_codename: COVID
period_length: month
```

Multi-value fields have space-separated values on a single line:

```
outcome_info.AESI: ADEM MYO ST
outcome_info.risk_upper: 42 28 42
```

Lines beginning with `#`, or lines that do not match any recognised key, are ignored. Section headers using the `---- Header ----` convention are also ignored. This means you can annotate your file freely without affecting how it is parsed.

The file must be named exactly **`options.txt`** and saved in the directory supplied to `options_file_location` in `aggregate_data()`.

---

## Complete field reference

### Study settings

#### `study_codename`

A short identifier for this study module. It is embedded in the output folder name (e.g. `RCA_COVID_SITE_20250101-20250131`) and in the `notes.txt` metadata file.

```
study_codename: COVID
```

#### `period_length`

The data from each cycle is automatically split into periods of this length. This length can be the same as the cycle length, or be an exact divisor of the cycle length if you would like to be able to see more granularity in your aggregated data. It does not effect the sequential analysis - each cycle is still treated as a single look at the data even if split into multiple periods - unless instructed to do so for the purpose of a retrospective-style analysis. 

It should be specified in the form: `month`, `week`, `day`, `2 months`, `2 weeks`, `2 days`, etc

```
period_length: month
```

#### `align_periods`

Whether to align periods to calendar boundaries (`TRUE`) or use fixed-length windows measured from the cycle start (`FALSE`).

With `TRUE` and `period_length: month`, the first period runs from the cycle start to the end of that calendar month, and subsequent periods follow calendar months exactly. This is useful for the first cycle, which aligns with the study start date and therefore is unlikely to perfectly align with the start of a month. This feature only works for a monthly period length.

With `FALSE`, each period is exactly one `period_length` in duration regardless of calendar boundaries.

```
align_periods: TRUE
```

#### `lookback_length`

The length of the outcome lookback period in **years**, measured back from the study start date. Outcomes occurring within this window are used to identify non-incident outcome that occur during the study period, which are then excluded from analysis.

This field is optional. If omitted, a default of 2 years is applied. If provided, it must be a positive number.

```
lookback_length: 2
```

> **Note:** Sites whose outcome data does not extend back the full `lookback_length` will receive a warning at validation and aggregation time, and the opportunity to specify the earliest date to which a lookback period can be applied. Aggregation can then continue. The date provided is stored in the notes.txt file to allow an appropriate caveat to be added during interpretation of results.

#### `ratio_precision`

The number of decimal places retained in ratio output columns. The recommended is at least 3.

```
ratio_precision: 3
```

#### `included_analyses`

A space-separated list of sub-analysis types to produce within each study design output, allowing stratification of results by vaccine type, subtype, dose number, or combinations thereof.

| Value | Vaccine stratification | Dose stratification |
|:---|:---|:---|
| `primary` | All vaccines pooled | All doses pooled |
| `subgroup_dose` | All vaccines pooled | By dose number (overall) |
| `subgroup_platform` | By vaccine type (`V_TYPE`) | All doses pooled |
| `subgroup_platform_dose` | By vaccine type (`V_TYPE`) | By dose number within type |
| `subgroup_brand` | By vaccine subtype (`V_SUBTYPE`) | All doses pooled |
| `subgroup_brand_dose` | By vaccine subtype (`V_SUBTYPE`) | By dose number within subtype |

All six values are typically included. Removing a value reduces output volume but also reduces the granularity of results available for analysis.

```
included_analyses: primary subgroup_dose subgroup_platform subgroup_platform_dose subgroup_brand subgroup_brand_dose
```

---

### Age groups

Age groups apply to all study designs, dividing the cohort into strata for stratified output tables. Age groups, along with sex, are critical for adjustment during analysis.

#### `age_groups.bounds`

Space-separated age boundaries in years. Must start with `0` and end with `Inf`. N boundary values define N−1 age groups.

#### `age_groups.labels`

Space-separated labels for each age group. The number of labels must equal the number of boundaries minus one (i.e. the number of groups).

```
age_groups.bounds: 0 5 10 15 20 25 30 35 40 45 50 55 60 65 70 75 80 Inf
age_groups.labels: 0-4 5-9 10-14 15-19 20-24 25-29 30-34 35-39 40-44 45-49 50-54 55-59 60-64 65-69 70-74 75-79 80+
```

---

### Outcome settings

The `outcome_info` fields are **parallel vectors** — each position corresponds to one AESI. All `outcome_info.*` fields must have the same number of values as `outcome_info.AESI`. All time values are in **days**.

#### `outcome_info.AESI`

Short codes for the Adverse Events of Special Interest included in this study. These codes must match the values found in `outcome_data$AESI`.

```
outcome_info.AESI: ADEM MYO ST
```

#### `outcome_info.clean_window`

The minimum number of days before an event required to classify it as incident. If a prior event of the same AESI occurred within this many days before the current event, the current event is excluded as a prevalent case.

```
outcome_info.clean_window: 365 183 365
```

#### `outcome_info.risk_lower` and `outcome_info.risk_upper`

The start and end of the post-vaccination risk window in days. Day 0 is the day of vaccination, day 1 the day after vaccination, and so on. Events occurring within `[risk_lower, risk_upper]` days of a vaccination are attributed to the vaccine.

```
outcome_info.risk_lower: 1  1  2
outcome_info.risk_upper: 42 28 42
```

#### `outcome_info.washout_post` and `outcome_info.washout_pre`

Days immediately after the risk window (`washout_post`) or before vaccination (`washout_pre`) that are excluded from the analysis. These buffers can be used to reduce misclassification bias and the effect of event-dependent exposures, respectively.

```
outcome_info.washout_post: 7 7 7
outcome_info.washout_pre:  7 7 7
```

#### `outcome_info.control_post_target` and `outcome_info.control_pre_target`

These are relevant only to the self controlled study designs. They determine the target ratio of control time to risk window length, for the post-vaccination and pre-vaccination control windows respectively. 

For example, a `control_post_target` of `1` means the algorithm attempts to allocate a control period equal in length to the risk window. A value of `3` means it attempts three times the risk window length. Actual lengths may vary in the presence of multiple vaccine doses or other censoring.

```
outcome_info.control_post_target: 1 1 1
outcome_info.control_pre_target:  3 3 3
```

#### `outcome_info.control_post_min` and `outcome_info.control_pre_min`

These are relevant only to the self controlled study designs. They determine the minimum target control period length in days. This overrides the target ratios above for AESI with very short risk windows. 

```
outcome_info.control_post_min: 30 30 30
outcome_info.control_pre_min:  30 30 30
```

---

### Vaccine settings

The `vaccine_info` fields are **parallel vectors** — each position pairs a subtype with its corresponding type. Both fields must have the same number of values.

#### `vaccine_info.vaccine_subtypes`

Vaccine subtype codes as they appear in `vaccination_data$V_SUBTYPE`. These represent the finest level of vaccine classification — typically a specific brand, formulation, or product name.

#### `vaccine_info.vaccine_types`

The broader vaccine type (platform) corresponding to each subtype above — for example, the delivery mechanism (mRNA, viral vector, protein subunit). Multiple subtypes may share the same type.

```
vaccine_info.vaccine_subtypes: ABC DEF GHI
vaccine_info.vaccine_types:    mRNA mRNA VVA
```

---

## Complete example

The following is a complete, valid `options.txt` file. Values are illustrative — in a multi-site study, use the file provided by the coordinating centre; in a single-site study, adapt this template to your study.

```
---- Analysis options ----
study_codename: COVID
lookback_length: 2
period_length: month
align_periods: TRUE
ratio_precision: 3
included_analyses: primary subgroup_dose subgroup_platform subgroup_platform_dose subgroup_brand subgroup_brand_dose
age_groups.bounds: 0 5 10 15 20 25 30 35 40 45 50 55 60 65 70 75 80 Inf
age_groups.labels: 0-4 5-9 10-14 15-19 20-24 25-29 30-34 35-39 40-44 45-49 50-54 55-59 60-64 65-69 70-74 75-79 80+
outcome_info.AESI: ADEM MYO ST
outcome_info.clean_window: 365 183 365
outcome_info.risk_lower: 1 1 2
outcome_info.risk_upper: 42 28 42
outcome_info.washout_post: 7 7 7
outcome_info.washout_pre: 7 7 7
outcome_info.control_post_target: 1 1 1
outcome_info.control_pre_target: 3 3 3
outcome_info.control_post_min: 30 30 30
outcome_info.control_pre_min: 30 30 30
vaccine_info.vaccine_subtypes: ABC DEF GHI
vaccine_info.vaccine_types: mRNA mRNA VVA
```

---

## Reading the file in R

`aggregate_data()` reads `options.txt` automatically when `options_file_location` is provided. You can also read and inspect the parsed options object directly:

```{r eval = FALSE}
options <- rapidcycler:::read_options_file("~/Documents/RCA/STUDY/SUBSTUDY")
str(options)
```

This returns a named list, which can be passed directly to `aggregate_data()` via the `options` argument (useful when you need to run multiple cycles without re-reading the file each time):

```{r eval = FALSE}
aggregate_data(
  ...,
  options = options,
  options_file_location = NULL  # required when passing options object directly
)
```

---

## Deprecated field names

Sites upgrading from earlier versions of *rapidcycler* may have `options.txt` files using older field names. These are still accepted with a warning:

| Deprecated key | Current key |
|:---|:---|
| `outcome_info.case_ascertainment` | `outcome_info.clean_window` |
| `vaccine_info.vaccine_brands` | `vaccine_info.vaccine_subtypes` |
| `vaccine_info.vaccine_platforms` | `vaccine_info.vaccine_types` |

Updating to the current names removes the warning.

---

## Troubleshooting

**`Options file (options.txt) not found at the location given.`**
The file does not exist at the path supplied to `options_file_location`. Check that the file is named exactly `options.txt` (not `options (1).txt` or similar) and that the directory path is correct.

**`Options object not in the correct format.`**
A required field is missing or has the wrong type. Check that all `outcome_info.*` fields have the same number of values as `outcome_info.AESI`, and that both `vaccine_info` fields have the same length.

**Warning: `The outcome_info.case_ascertainment key in options.txt is deprecated.`**
Rename `outcome_info.case_ascertainment` to `outcome_info.clean_window` in your file.

**Warning: `The vaccine_info.vaccine_brands key in options.txt is deprecated.`**
Rename `vaccine_info.vaccine_brands` to `vaccine_info.vaccine_subtypes` in your file.
