---
title: "Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{validate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The purpose of validation is to ensure that the input datasets required for aggregation are in the correct format. The primary function for validation is `validate_input_data()`. This function is called automatically when you run `aggregate_data()`, so it is not necessary to perform validation manually. However, running validation separately before aggregation is useful for catching data issues early, especially when working with large datasets.

### What is checked

Validation covers four main areas:

**Person data**

- All required columns are present (`PID`, `DOB`, `SEX`) and optional columns (`ENROL_DATE`, `CENSOR_DATE`, `CENSOR_TYPE`) are correctly typed if present
- `PID` values are unique (no duplicate persons)
- Date columns (`DOB`, `ENROL_DATE`, `CENSOR_DATE`) are stored as 8-digit integers in `yyyymmdd` format
- Date consistency: `DOB` ≤ `ENROL_DATE` ≤ `CENSOR_DATE`
- `CENSOR_TYPE` values are from the allowed set (`"missing"`, `"dead"`, `"travelled"`, `"emigrated"`)
- At least some persons are active during the cycle period

**Vaccination data** (skipped if `vaccination_data = NULL`)

- All required columns are present (`PID`, `V_DATE`, `V_SUBTYPE`) and correctly typed
- All `PID` values exist in `person_data`
- `V_DATE` is in valid 8-digit integer format
- `V_SUBTYPE` codes match those defined in the options file
- Vaccination dates fall within each person's active period (after `ENROL_DATE`, before `CENSOR_DATE`)

**Outcome data**

- All required columns are present (`PID`, `AESI`, `EVENT_DATE`, `ENCOUNTER_TYPE`) and correctly typed
- All `PID` values exist in `person_data`
- `AESI` codes match those defined in the options file
- `EVENT_DATE` is in valid 8-digit integer format
- Event dates fall within each person's active period

**Lookback coverage**

Checks that outcome data extends far enough back before the first vaccination (or cycle start, for historical aggregation) to satisfy the lookback requirement. By default, at least 75% of the configured lookback period (default: 2 years) must be covered. If insufficient coverage is detected, an interactive prompt allows you to provide a custom lookback start date to confirm awareness of the limitation.

### Return value

`validate_input_data()` returns a character vector of issue descriptions. An empty vector means no issues were found and the data is ready for aggregation. When issues are present, each entry describes a specific problem to address.

```{r, eval=FALSE}
issues <- validate_input_data(person_data, vaccination_data, outcome_data,
                              cycle_start_date = as.Date("2022-01-01"),
                              cycle_end_date = as.Date("2022-01-31"))

if (length(issues) == 0) {
  message("No issues found.")
} else {
  print(issues)
}
```

### Options file

`validate_input_data()` reads the same `options.txt` file as `aggregate_data()` — it uses this to check that vaccine subtype codes and AESI codes in the input data match those defined for the study. The `options_file_location` parameter specifies the folder containing `options.txt` and defaults to the current working directory (`getwd()`). You can also supply a pre-parsed options object directly via the `options` parameter (see `?read_options_file`).

### Set up for examples

Load the package:

```{r setup, message=FALSE}
library(rapidcycler)
library(dplyr)
library(magrittr)
```

If you would like to use synthetic data to run the examples, you can create and save that in the current working directory with the code below:

```{r, eval=FALSE}
generate_synthetic_data(pop_size = 1000000)
```

Or, if you already have some data saved in the current working directory, load it as so:

```{r, eval=FALSE}
person_data <- data.table::fread("person_data.csv")
vaccination_data <- data.table::fread("vaccination_data.csv")
outcome_data <- data.table::fread("outcome_data.csv")
```

### Example: Valid input data

```{r, eval=FALSE}
# options.txt is assumed to be in the current working directory
validate_input_data(person_data, vaccination_data, outcome_data,
                    cycle_start_date = as.Date("2022-01-01"),
                    cycle_end_date = as.Date("2022-01-31"))
# Returns character(0) — no issues found
```

To specify a different location for the options file:

```{r, eval=FALSE}
validate_input_data(person_data, vaccination_data, outcome_data,
                    cycle_start_date = as.Date("2022-01-01"),
                    cycle_end_date = as.Date("2022-01-31"),
                    options_file_location = "~/Documents/RCA_COVID/")
```

### Example: Invalid input data

The example below introduces three deliberate errors: an incorrect column type for `DOB`, a renamed vaccination date column, and an unrecognised AESI code. The returned vector will describe each issue.

```{r, eval=FALSE}
issues <- validate_input_data(
  person_data %>% mutate(DOB = as.character(DOB)),       # DOB should be integer, not character
  vaccination_data %>% rename(VAC_DATE = V_DATE),        # V_DATE column missing
  outcome_data %>% mutate(AESI = if_else(AESI == "ADEM", "ADM", AESI)), # unrecognised AESI code
  cycle_start_date = as.Date("2022-01-01"),
  cycle_end_date = as.Date("2022-01-31")
)
print(issues)
```

### Example: Historical input data

When `vaccination_data = NULL`, vaccination-related checks are skipped. This is used when aggregating data from a period before the vaccination campaign began.

```{r, eval=FALSE}
validate_input_data(person_data,
                    vaccination_data = NULL,
                    outcome_data,
                    cycle_start_date = as.Date("2022-01-01"),
                    cycle_end_date = as.Date("2022-01-31"))
```
